---
title: "Making plots informative"
subtitle: "<br>EC 103--02"
author: "Marcio Santetti"
footer: "EC 103--02"
format: 
  revealjs:
    theme: lab.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
knitr:
  opts_chunk:
#    dev: "ragg_png"
    retina: 3
    dpi: 200
execute:
  freeze: auto
  fig-width: 5
  fig-height: 6
---

```{r}
#| include: false
# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

# What is your plot about?

## What is your plot about?

<br>

In the previous lab, we learned how to map data into *aesthetics*, using the "grammar of graphics" with the `ggplot2` package.

. . .

<br>


However, that is not enough!

<br>

. . .

Always keep in mind that *any* person should be able to **read** and **interpret** what your figure wants to express.



## What is your plot about?

<br>

```{r}
#| echo: false


library(tidyverse)
library(ggeasy)

unemp_data <- read_csv("https://raw.githubusercontent.com/marciosantetti/ec103-fall22/main/lectures/005-unemp/unemp-data.csv")
unemp_data %>% 
  ggplot(aes(y = unemployed, x = period)) +
  geom_line() +
  labs(x = "", y = "u") +
  theme_gray(base_size=16)
```

## What is your plot about?


```{r}
#| echo: false
#|message: false

unemp_data %>% 
  ggplot(aes(y = unemployed, x = period)) +
  geom_line() +
  labs(x = "", y = "u") +
  labs(title = "A title") +
  theme_gray(base_size=16)
```



## What is your plot about?


```{r}
#| echo: false
#|message: false

unemp_data %>% 
  ggplot(aes(y = unemployed, x = period)) +
  geom_line() +
  labs(x = "", y = "u") +
  labs(title = "A title",
       subtitle = "Maybe a subtitle?") +
  theme_gray(base_size=16)
```


## What is your plot about?


```{r}
#| echo: false
#|message: false

unemp_data %>% 
  ggplot(aes(y = unemployed, x = period)) +
  geom_line() +
  labs(x = "", y = "u") +
  labs(title = "A title",
       subtitle = "Maybe a subtitle?",
       x = "What variable is here?") +
  theme_gray(base_size=16)
```



## What is your plot about?


```{r}
#| echo: false
#|message: false

unemp_data %>% 
  ggplot(aes(y = unemployed, x = period)) +
  geom_line() +
  labs(x = "", y = "u") +
  labs(title = "A title",
       subtitle = "Maybe a subtitle?",
       x = "What variable is here?",
       y = "What variable is here?") +
  theme_gray(base_size=16)
```



## What is your plot about?


```{r}
#| echo: false
#|message: false

unemp_data %>% 
  ggplot(aes(y = unemployed, x = period)) +
  geom_line() +
  labs(x = "", y = "u") +
  labs(title = "A title",
       subtitle = "Maybe a subtitle?",
       x = "What variable is here?",
       y = "What variable is here?",
       caption = "What is the data source?") +
  theme_gray(base_size=16)
```


## What is your plot about?


<br><br>

In a plot, **labels** matter.

. . .

<br>

And also do **scales**.



## What is your plot about?


```{r}
#| echo: false
#|message: false


library(scales)

unemp_data %>% 
  ggplot(aes(y = unemployed, x = period)) +
  geom_line() +
  labs(x = "", y = "") +
  labs(title = "A title") +
  theme_gray(base_size=16) +
  scale_y_continuous(labels = comma)
  
```


## What is your plot about?


```{r}
#| echo: false
#| message: false
#| dev: "svg"


library(scales)

unemp_data %>% 
  mutate(un_rate = unemployed / labor_force) %>% 
  ggplot(aes(y = un_rate, x = period)) +
  geom_line() +
  labs(x = "", y = "") +
  labs(title = "A title") +
  theme_gray(base_size=16) +
  scale_y_continuous(labels = percent_format(scale = 100))

  
```

## What is your plot about?


```{r}
#| echo: false
#| message: false
#| dev: "svg"


library(scales)

unemp_data %>% 
  mutate(un_rate = unemployed / labor_force) %>% 
  ggplot(aes(y = un_rate, x = period)) +
  geom_line() +
  labs(x = "", y = "") +
  labs(title = "A title") +
  theme_gray(base_size=16) +
  scale_y_continuous(breaks = seq(0.02, 0.10, by = 0.01),
                     labels = percent_format(scale = 100))

  
```


## What is your plot about?


```{r}
#| echo: false
#| message: false
#| dev: "svg"


library(scales)

unemp_data %>% 
  mutate(un_rate = unemployed / labor_force) %>% 
  ggplot(aes(y = un_rate, x = period)) +
  geom_line() +
  labs(x = "", y = "") +
  labs(title = "A title") +
  theme_gray(base_size=16) +
  scale_y_continuous(breaks = seq(0.02, 0.10, by = 0.01),
                     labels = percent_format(scale = 100)) +
  scale_x_continuous(breaks = seq(from = 1950, to = 2021, by = 10))

  
```

# RStudio time!

