---
title: "Customizing plots"
subtitle: "<br>EC 103--02"
author: "Marcio Santetti"
footer: "EC 103--02"
format: 
  revealjs:
    theme: lab.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
knitr:
  opts_chunk:
#    dev: "ragg_png"
    retina: 3
    dpi: 200
execute:
  freeze: auto
  fig-width: 5
  fig-height: 6
---

```{r}
#| include: false
# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```


# Themes

## Themes

<br>

Some useful packages:


- [`{ggthemes}`](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/)

- [`{hrbrthemes}`](https://github.com/hrbrmstr/hrbrthemes)

- [`Nice blog post by David Keyes`](https://rfortherestofus.com/2019/08/themes-to-improve-your-ggplot-figures/)

# Fonts

## Fonts

<br>

The easiest way to use custom fonts is through the {showtext} package.

. . .

<br>

```{r}
#| echo: true
#| message: false

library(showtext) # Make sure to install it first!


```

. . .

<br>

And importing fonts from [`Google Fonts`](https://fonts.google.com/).







# Colors

## Colors

<br>

R has its own built-in colors:

[`R colors`](https://www.datanovia.com/en/blog/awesome-list-of-657-r-color-names/)

. . .

<br>

However, you can use any color you want through *hex codes*.

[`https://www.color-hex.com/`](https://www.color-hex.com/)


# Patchworking

## Patchworking

<br><br>


```{r}
#| echo: true
#| message: false

library(patchwork) # Make sure to install it first!


```



# Saving your plots

## Saving your plots

<br><br>


R allows you to save your plots either as an image or as a PDF file.