---
title: ".b[Economic Growth]"
subtitle: ".b[.green[EC 103--002]]"
author: "Marcio Santetti"
date: "Fall 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork, readxl, MetBrewer)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())

```



# Motivation



---

# Housekeeping

<br><br>


* .b[Required reading]:

  * [`OpenStax, ch. 7`](https://openstax.org/books/principles-macroeconomics-2e/pages/7-introduction-to-economic-growth).


---

# Growth: A recent phenomenon

<br><br><br>

In human history, economic growth is a relatively .hi[recent] phenomenon.

--

<br>

Now that we have studied what is the main measure of economic performance, we can come back to the *"hockey stick"* chart from previous lectures.



---

# Growth: A recent phenomenon


```{r, echo = FALSE, message = FALSE, dev = "svg", fig.height = 5.5}

dat <- read_xlsx("maddison.xlsx", sheet = 3)

dat_sample <- dat %>% 
  filter(country %in% c("United Kingdom", "United States", "Republic of Korea", "China")) %>% 
  filter(year >= 1700)

dat_sample %>% 
  ggplot(aes(x = year, y = gdppc, color = country)) +
  geom_line(size = 0.8) +
  scale_y_continuous(labels=scales::dollar_format()) +
  scale_color_met_d("Homer2") +
  labs(x = "", y = "GDP per capita", title = "Gross domestic product per capita, selected countries (1700—2018)") +
  easy_legend_at("bottom") +
  easy_remove_legend_title() +
  easy_plot_legend_size(13) +
  easy_y_axis_title_size(13)
```


---

# Growth: A recent phenomenon

Measuring .hi-orange[GDP *per capita*] (i.e., *GDP/total population*) is a valid (but rough) measure of a nation's overall standard of living, as it measures much output each individual would receive in case total output were evenly divided across society.

--

If, instead, we turn our attention to economic growth, the concept of .hi[output per worker] seems more appealing.

--

  - Think about for .hi-orange[how long] your parents/grandparents had to work in order to afford basic goods, such as food and clothing.
  
  - It is very likely that you will have to work relatively .hi[less time] to afford the same things&mdash;and probably with a higher *quality*!
  
--

The .hi[main reason] why this happens is because .hi-orange[output per worker] has been increasing over time.



---

# Growth: A recent phenomenon


.hi-orange[Output per worker] measures how much output each worker, on average, is producing.

  - It is .hi[not] the same as output per worker, since not everyone in the population *formally* works.
  
  - Also, output *per capita* may *fall* while output per worker is *increasing*!
  
--

<br>

Given its definition, another way to call output per worker is .hi[labor productivity].
  
--

<br>

  - But .hi[why] has labor productivity increased?
  
  - And what .hi-orange[factors] allow for a *sustained* growth in output per worker?

---

# Growth: A recent phenomenon

```{r, echo = FALSE, message = FALSE, dev = "svg", fig.height = 5.5}

data <- read_csv("epwt.csv")

data %>% 
  filter(Country %in% c("United States", "Korea Republicof", "China", "South Africa", "Australia")) %>% 
  ggplot(aes(y = x, x = Year, color = Country)) +
  geom_line() +
  geom_point() +
  scale_color_met_d("Hokusai3") +
  scale_y_continuous(labels=scales::dollar_format()) +
  labs(y = "Output per worker ($ 2005 PPP/worker)",
       x = "",
       title = "Labor productivity over time (1970—2008)") +
  easy_plot_legend_size(13) +
  easy_y_axis_title_size(13)
```


---
layout: false
class: inverse, middle

# Components of economic growth

---

# Components of economic growth


We all understand that a *steady* increase in GDP over time implies an economic growth path.

--

<br>

.right[
But what makes the economy .hi[*actually*] grow?
]

--

<br>

Among others, we may highlight a few important .hi-orange[components] of economic growth:

1. Physical capital (infrastructure);

2. Human capital;

3. Technology.


---

# Components of economic growth

<br>

.hi[Physical capital] includes the overall infrastructure firms and government use to create and distribute their production.

  * This infrastructure ranges from firm plants, physical equipment (e.g., machinery), roads, ports, etc.
  
--

<br>

Not only can a larger and better infrastructure increase the *quantity* of output generated, but also its *quality*.

--

<br>

.right[Examples?]

---

# Components of economic growth

.hi[Human capital] embraces the *skills* and knowledge that allow workers to increase their dexterity and, consequently, their productivity.

--

<br>

If the amount of physical capital is held *fixed*, a more .hi[skilled] worker can produce *more* output using the same structures relative to a *less* skilled one.

--

<br>

[`Our World in Data: Literacy`](https://ourworldindata.org/literacy)

---

# Components of economic growth

.hi[Technological improvement] is way more than inventing newer machinery, or creating better smartphones.

--

New forms of organizing production are .hi[also] technological improvements.

<br>

--

When the .hi[quality] of machinery and other production equipment increase, we call it .hi-orange[embodied technical change].
  
  - Faster computer chips, fuel-efficient cars,...

--


Even if the quality and quantity of physical capital remain the same, labor productivity may still increase due to .hi[changes] in the production process.
  
  - This is known as .hi-orange[disembodied technical change].
  
---

# Components of economic growth

Technical change generally takes place in .hi[two] stages:

  - Invention;
  
  - Innovation.
  
--

An .hi-orange[invention] involves knowledge advancements.

--

When this new knowledge is put into production&mdash;to develop new products or to make existing processes more efficient&mdash;we have an .hi[innovation].

--

A great way of looking at how a country values investing is research is to verify the .hi[share] of GDP spent on Research & Development.

  - [`Data from OECD`](https://data.oecd.org/rd/gross-domestic-spending-on-r-d.htm)


---

layout: false
class: inverse, middle

# Calculating economic growth


---

# Calculating economic growth

.smaller[Starting from:]

```{r, echo = FALSE, message = FALSE, dev = "svg", fig.height = 5}

dat %>% 
  filter(country %in% c("United States")) %>% 
  filter(year >= 1940) %>% 
  ggplot(aes(x = year, y = gdppc)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels=scales::dollar_format()) +
  labs(x = "", y = "GDP per capita",
       title = "Gross domestic product per capita, US (1940—2018)") +
  easy_y_axis_title_size(13)

```



---

# Calculating economic growth


.pull-left[
```{r, echo = FALSE, message = FALSE, dev = "svg", fig.asp = 0.8}

dat %>% 
  filter(country %in% c("United States")) %>% 
  filter(year >= 1940) %>% 
  ggplot(aes(x = year, y = gdppc)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels=scales::dollar_format()) +
  labs(x = "", y = "GDP per capita",
       title = "Gross domestic product per capita, US (1940—2018)") +
  easy_y_axis_title_size(13)

```

]

.pull-right[
Looking at this measure in .hi[levels] shows the overall long-run process of economic growth.

But if we would like to know what was the .hi-orange[growth rate] over time, we need to compute these rates from the data in levels.

.smaller[
$$
\begin{aligned}
\text{Growth rate (%)} = \dfrac{\text{Final Period - Initial Period}}{\text{Initial Period}} \ \times 100
\end{aligned}
$$
]


]

---

# Calculating economic growth

.smaller[Then, from the first chart, we can calculate the year-to-year growth rate in GDP per capita:]

```{r, echo = FALSE, message = FALSE, dev = "svg", fig.height = 5}

dat_sample2 <- dat %>% 
  filter(country %in% c("United States")) %>% 
  filter(year >= 1940)


dat_sample2 %>% 
  mutate(gdppc_growth = ( ( gdppc - lag(gdppc, n = 1) ) / lag(gdppc, n = 1) ) * 100 ) %>% 
  ggplot(aes(x = year, y = gdppc_growth)) +
  geom_line() +
  geom_point() + 
  scale_y_continuous(labels=scales::percent_format(scale = 1)) +
  labs(x = "", y = "GDP per capita",
       title = "Growth rate: Gross domestic product per capita, US (1940—2018)") +
  easy_y_axis_title_size(13)
```

---

# Calculating economic growth

From our previous lecture,

```{r, echo=F}
real_gdp <- tibble(
  "Quarter" = c("2008q1", "2008q2", "2008q3", "2008q4", "2009q1", "2009q2", "2009q3", "2009q4"),
  "Nominal GDP" = c(14373.9, 14497.8, 14546.7, 14347.3, 14178.0, 14151.2, 14242.1, 14453.8), 
  "Real GDP" = c(13366.9, 13415.3, 13324.6, 13141.9, 12925.4, 12901.5, 12973.0, 13149.5)
)

real_gdp <- real_gdp %>% 
  mutate("GDP deflator" = `Nominal GDP`/`Real GDP`)



real_gdp %>% 
  select(-`Real GDP`) %>% 
  kable(digits = 2)

```

--

<br>

Calculate the .hi[growth rate] of nominal GDP between the first and the last periods.




---

layout: false
class: inverse, middle

# Next time: Growth discussions


---
exclude: true