---
title: "Introduction to R, RStudio, and the `tidyverse`"
subtitle: "EC 103--02"
author: "Marcio Santetti"
footer: "EC 103--02"
format: 
  revealjs:
    theme: lab.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
---


```{r}
#| include: false
# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

# Welcome to our `Lab Sessions`!

# **What is R?**

## What is R?


`R` is an open source (i.e., *free!*) statistical/programming language that serves **multiple purposes**.

. . .

The purpose of using `R` in our course is to *manipulate* and *visualize* economic data.

. . .

<br>

If this will be your first experience with **coding**, welcome!

  - We'll help each other out!
  


## What is R?

`R` works by the users typing **functions**, which come from **packages**

. . .

Packages only need to be installed **once** *per machine*.

<br>

. . .

This is done by using the `install.packages()` function:

<br>

:::small
> `install.packages("name of the desired package")`
:::


## What is R?

And whenever we wish to use functions from a certain package, we first call the package using the `library()` function:

<br>

:::small
> `library(name of the desired package)`
:::



```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(hrbrthemes)

theme_set(theme_ipsum_rc())
```


# What is RStudio?

## What is RStudio?

`R`Studio is the **IDE** (Integrated Development Environment) we will use to run the `R` language.

. . . 

It is a very convenient space, where we can write our nscripts, look up our data, and visualize it!

. . .

<br><br><br>

Now off to a little tour!

# What is the `tidyverse`?

## What is the `tidyverse`?

<br>

The `tidyverse` is a *metapackage* (i.e., a set of packages) that allows us to import, manipulate, and visualize data.

<br><br><br>


<https://www.tidyverse.org/>

## What is the `tidyverse`?



<br><br><br>


Let us install it properly and play around with some of its functions now.









