---
title: "Variable classes"
subtitle: "EC 103--02"
author: "Marcio Santetti"
footer: "EC 103--02"
format: 
  revealjs:
    theme: lab.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
---

```{r}
#| include: false
# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

# Dealing with variable classes


## Dealing with variable classes

<br><br>

Every column in our data sets pertains to a certain **class**.



## Dealing with variable classes


```{r}
#| echo: true
#| message: false


library(tidyverse)
unemployment_rate <- read_csv("unrate.csv")

unemployment_rate
```


## Dealing with variable classes

<br><br>

But sometimes our variables are imported with a **different class** than the one we desire.


## Dealing with variable classes


```{r}
#| echo: true
#| message: false


net_exports <- read_csv("netexp.csv")

net_exports

```


## Dealing with variable classes

```{r}
#| echo: true
#| message: false


net_exports %>% 
  ggplot(aes(x = date, y = netexp)) +
  geom_line()


```


## Dealing with variable classes

```{r}
#| echo: false
#| message: false


library(tsibble)

net_exports %>% 
  mutate(period = seq(as.Date("1947-01-01"), by = "quarter", length.out = 302)) %>% 
  ggplot(aes(x = period, y = netexp)) +
  geom_line() +
  scale_x_yearquarter()


```



## Dealing with variable classes


<br><br>


Let us learn how to deal with variable classes in practice.