---
title: ".b[Unemployment, pt. II]"
subtitle: ".b[.green[EC 103--002]]"
author: "Marcio Santetti"
date: "Fall 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork, janitor)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())

```



# Motivation



---

# Housekeeping

<br>

* .b[Required reading]:

  * [`CORE, ch. 13, section 13.2`](https://www.core-econ.org/the-economy/book/text/13.html#132-output-growth-and-changes-in-unemployment)
  
  
---

# Last time

<br>

Last time, we looked at some official US .hi[data] on unemployment, as well as some basic .hi-orange[definitions].

--


<br><br>

Now it is time to investigate some of the most relevant .hi[relationships] that unemployment shares with other .hi-orange[macroeconomic] variables.

--

<br><br>

Lastly, we will also study some of the main .hi[explanations for] unemployment, since it may come from different phenomena.




---


layout: false
class: inverse, middle

# Unemployment and other aggregates

---


# Unemployment and other aggregates


When output grows *quickly*, unemployment tends to .hi[fall].

--

When output grows *more slowly* or *falls*, unemployment tends to .hi[rise].

--

  - Why would this happen?
  
  - .smaller[Lower unemployment &rarr; higher wages &rarr; increased workers' bargaining power &rarr; better income distribution.]

--

<br>

[`How do the data look?`](https://fred.stlouisfed.org/series/GDPC1)

--

<br>


This association is known as .hi-orange[Okun's law].

---

# Unemployment and other aggregates

<br>

In more detail, Okun's law claims that there is a stable relationship between the .hi[change in the unemployment rate] $(\Delta u)$ and .hi-orange[real GDP growth] $(g_y)$.

<br>

$$
\begin{aligned}
\Delta u = a + b \times (g_y)
\end{aligned}
$$

<br>

where $b$ is called the .hi[Okun coefficient], and $a$ is the change in unemployment when the economy does not grow in a given period (i.e., $g_y = 0$).

---


# Unemployment and other aggregates

```{r, echo=FALSE, dev = "svg", fig.height = 5.5, warning=FALSE}
dat <- read_csv("unrate_annual.csv")


dat <- dat %>% 
  clean_names() %>% 
  mutate(delta_u = c(NA, diff(unrate)),
         gdp_growth = ( ( gdp - lag(gdp, n = 1) ) / lag(gdp, n = 1) ) * 100 )


dat %>% 
  ggplot(aes(x = gdp_growth, y = delta_u)) + 
  geom_point(size = 2, color = "#800000", shape = 5) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  labs(x = "GDP growth (%)",
       y = "Change in unemployment (p.p.)",
       title = "Okun's law: United States, 1949–2021") +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)

```


---


# Unemployment and other aggregates

```{r, echo=FALSE, dev = "svg", fig.height = 5.5, warning=FALSE}

dat %>% 
  mutate(years = as.character(period)) %>% 
  filter(years %in% c("1970", "2008", "2019", "2000", "2007", "2021")) %>% 
  ggplot(aes(x = gdp_growth, y = delta_u)) + 
  geom_point(size = 3, color = "#800000", shape = 5) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  labs(x = "GDP growth (%)",
       y = "Change in unemployment (p.p.)",
       title = "Okun's law: United States, selected years") +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13) +
  geom_text_repel(aes(label = years))



```


---


# Unemployment and other aggregates

Based on these data, the *Okun coefficient* (*b*) equals -0.4, and *a*, 1.33. 

--

  * What does this mean?

--

<br>

And what if one wants to keep unemployment constant (meaning $\Delta u = 0$)? How much would the economy have to grow?

---

# Unemployment and other aggregates

It would take this amount of real GDP growth just to keep unemployment constant $(\Delta u = 0)$ for two main reasons:

  * .hi[Population] (and, consequently, the labor force) is still growing;
  
  * .hi[Labor productivity] is rising over time (implying that output per worker grows faster than employment).
  
--

In other words, each year there are more individuals *looking for jobs* (due to population growth), while technology keeps improving, so that *less people* are needed to produce the same amount of output over time (productivity growth).

--

Therefore, GDP has to grow to keep the change in unemployment .hi[steady] over time.

---



# Unemployment and other aggregates

How does unemployment relate with the .hi[price level]?

--

In other words, when *less* people are unemployed, does this affect .hi-orange[inflation]?


--

<br><br>

This will be next week's topic.

---


layout: false
class: inverse, middle

# Explaining unemployment


---

# Explaining unemployment

<br>

The reasons for high unemployment are .hi[manifold].

--

<br>

Some of the main .hi-orange[causes] of unemployment may be classified by:

  * .hi[Demand];
  
  * .hi[Structural];
  
  * .hi[Frictional] factors.



---

# Explaining unemployment


Some people may be unemployed because .hi[not enough] goods or services that *require* their labor are being currently produced.

--

  * Thus there is a .hi[lack of demand] for these workers.


--

When the available jobs in the economy are .hi[not matched] by the current candidates (i.e., there is a a *mismatch* between jobs and workers), unemployed is classified as .hi-orange[structural].

--

  * As potential reasons, there may be a lack of the necessary skills for the job, or perhaps the necessary workers are located in different parts of the country.
  
--

Throughout a person's worklife, they may be *transitioning* between jobs, moving to a new area, entering the job market after graduating, etc.

--

  * This kind of unemployment is considered .hi[frictional], comprehending just a "looking" period.

---

# Explaining unemployment

<br>

Different .hi-orange[causes], different .hi[solutions].

--

Depending on what we believe the cause of current unemployment is, we may work with different .hi[policies/solutions].

--


<br><br>

.right[
What are some .hi-orange[procedures] for each cause of unemployment?
]

---

# Explaining unemployment


Whether unemployment is demand-driven or structural, it is a matter of policy discussion.

--

However, looking at the available .hi[data] always helps.

--

<br>

The .hi[Beveridge curve] is a relationship that puts together the number of .hi-orange[unemployed] workers and the number of .hi-orange[job vacancies]. 

--

  * In a scenario with *low* vacancies but *high* unemployment, the latter is probably due to the .hi[lack of demand].

--

  * Conversely, with *low* unemployment and a *high* number of job vacancies, unemployment will likely be .hi[structural].

---

# Explaining unemployment

```{r, echo=FALSE, dev = "svg", fig.height = 5.5, warning=FALSE}
unrate <- read_csv("unrate.csv")
open <- read_csv("job-openings-rate.csv")


unrate <- unrate %>% 
  filter(DATE >= "2000-12-01" & DATE <= "2022-06-01")


unrate <- unrate %>% select(UNRATE)

open <- open %>% 
  add_column(unrate) %>% 
  rename(openings = JTSJOR,
         unrate = UNRATE)


open %>% 
  filter(DATE >= "2020-05-01") %>% 
  ggplot(aes(y = openings, x = unrate)) +
  geom_point(alpha = 0.8, size = 1.7) +
  labs(title = "The US Beveridge Curve", 
       subtitle = "May 2020 - June 2022",
       x = "Unemployment rate",
       y = "Job openings rate") +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)



```



---


layout: false
class: inverse, middle

# Next time: Inflation `r emo::ji("ghost")`


---
exclude: true