---
title: ".b[Unemployment, pt. I]"
subtitle: ".b[.green[EC 103--002]]"
author: "Marcio Santetti"
date: "Fall 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())

```



# Motivation



---

# Housekeeping

<br>

.b[Required readings]:

  * [`OpenStax, ch. 8`](https://openstax.org/books/principles-macroeconomics-2e/pages/8-introduction-to-unemployment)
    
      * Sections 8.1 and 8.2.
      
<br>

  * [`How the Government Measures Unemployment (BLS)`](https://www.bls.gov/cps/cps_htgm.htm)
  
  
---

# A look at the data

<br><br><br>

[`Offical US unemployment data`](https://fred.stlouisfed.org/series/UNRATE)



---

# A plural phenomenon

<br>

Unemployment is much .hi[more] than an *economic* statistic.

--

Not only do its consequences affect one's .hi-orange[financial] provision, but also .hi[familiar] relations, and .hi-orange[mental] and .hi[physical] health.

--

<br>

An .hi[analogy] to understand unemployment is to compare it with a functional, but unused *factory*.

--

Thus unemployment is a .hi-orange[signaling] measure that output (i.e., GDP) may be falling short of .hi[potential] output&mdash;that is, the *maximum* the economy can produce without rising inflation or completely depleting raw materials.


---

# A plural phenomenon

<br><br><br>

[`Potential GDP for the US economy`](https://fred.stlouisfed.org/series/GDPPOT)


---

layout: false
class: inverse, middle

# Basic definitions

---

# Basic definitions


In the US, the .hi[Bureau of Labor Statistics] tracks and reports all unemployment-related data over time.

  * [`BLS website`](https://www.bls.gov/bls/unemployment.htm)
  
--

<br>

Defining who is unemployed is more .hi-orange[complex] than just calculating the share of the total population that is currently not working.

--

<br>

  * A person is considered .hi[unemployed] if they do not have a job, but are currently available to work, and actively looking for work in the previous .green[four] weeks;
  
  * A person is .hi-orange[employed] if currently working for pay.


---

# Basic definitions

Every month, the US government carries out the .hi-orange[Current Population Survey] (CPS), off of which the unemployment rate for the previous month is based.

--

  - It began in 1940, and in 1942 the US Census Bureau took over the CPS.
  
--

There are about .hi[60,000] eligible sample households, and the Bureau employees ask several labor-market-related questions regarding the reference week (which usually includes the 12th day of the month). 

--

<br>

Every *first* Friday of each month, the unemployment rate for the previous month is announced.

  * [`See the latest numbers`](https://www.bls.gov/cps/home.htm)
  
---

# Basic definitions

<br><br>

Despite being incredibly complex and representative, measuring the unemployment rate is *not* free from .hi[criticism].

  
  * What about care work? Cleaning? House chores? (Both paid and unpaid)


  

---

# Basic definitions

For unemployment computing purposes, a useful concept is the .hi[labor force].

--

The labor force includes all individuals that are .hi-orange[currently] employed *and* the .hi[unemployed] (i.e., actively looking for a job).

--

  * To be considered .hi[out of the labor force], one has to have either .hi-orange[given up] looking for work *or* or not been currently available to work (due to a parental leave, injury, etc.).
  * Also, full-time students, retirees, institutionalized individuals, those staying home to take care of children, and discouraged job seekers are considered .hi[out] of the labor force.


--


Furthermore, unemployment statistics only include the .hi-orange[adult] population (16-and-over).

--

Thus, the .hi[unemployment rate] is the percentage (%) of adults who are in the labor force but who .hi-orange[do not] have jobs:

$$
\begin{aligned}
\text{Unemployment rate} = \dfrac{\text{Unemployed}}{\text{Total labor force}} 
\end{aligned}
$$

---

# Basic definitions

```{r, echo=FALSE, message=FALSE, dev = "svg", fig.height=5.5}
un_data <- read_csv("unemp-data.csv")

cols1 <- c("Employed" = "#30845e", "Unemployed" = "#c76161")


un_data <- un_data %>% 
  mutate(emp_pop_ratio = (employed / non_inst_pop) * 100,
         unemp_rate = (unemployed / labor_force) * 100,
         pct_lf = (labor_force / non_inst_pop) * 100)

un_data %>% 
  ggplot() +
  geom_line(aes(x = period, y = employed, color = "Employed"), size = 0.8) +
  geom_line(aes(x = period, y = unemployed, color = "Unemployed"), size = 0.8) +
  labs(y = "Number of persons", x = "", color = "Measure",
       title = "Labor market data: US, 1950–2021") +
  easy_legend_at("bottom") +
  easy_y_axis_title_size(13) +
  easy_plot_legend_size(13) +
  scale_color_manual(values = cols1) +
  scale_y_continuous(labels = scales::comma)

```

---

# Basic definitions

```{r, echo=FALSE, message=FALSE, dev = "svg", fig.height=5.5}

cols <- c("Employed" = "#30845e", "Unemployed" = "#c76161", "Labor force" = "#528ff5")

un_data %>% 
  ggplot() +
  geom_line(aes(x = period, y = employed, color = "Employed"), size = 0.8) +
  geom_line(aes(x = period, y = unemployed, color = "Unemployed"), size = 0.8) +
  geom_line(aes(x = period, y = employed + unemployed, color = "Labor force"), size = 0.8) +
  labs(y = "Number of persons", x = "", color = "Measure",
       title = "Labor market data: US, 1950–2021") +
  easy_legend_at("bottom") +
  easy_y_axis_title_size(13) +
  easy_plot_legend_size(13) +
  scale_color_manual(values = cols) +
  scale_y_continuous(labels = scales::comma)


```




---

# Basic definitions

<br><br>


Along with the unemployment rate, another important labor market statistic is the labor force .hi[participation rate].

--

This rate is obtained by *dividing* the total labor force by the adult population:

<br>

$$
\begin{aligned}
\text{LF participation rate} = \dfrac{\text{Labor force}}{\text{Total adult population}} 
\end{aligned}
$$

---

# Basic definitions

<br><br>

Finally, we have the .hi[employment-to-population ratio].

--

As the name suggests, it is the .hi-orange[ratio] of individuals with paid work&mdash;of any kind&mdash;to the total population.

--

<br>

This measure is convenient when the goal is to .hi[compare] (un)employment statistics across countries.

  * Also keep in mind that each country has its *own* methodology to compute unemployment statistics.

---

# Basic definitions

```{r, echo=FALSE, message=FALSE, dev = "svg", fig.height=5.5}
colors = c("Employment to population ratio" = "#30845e", "Labor force participation rate" = "#c76161")

un_data %>% 
  ggplot(aes(x = period)) +
  geom_point(aes(y = emp_pop_ratio, color = "Employment to population ratio")) +
  geom_line(aes(y = emp_pop_ratio, color = "Employment to population ratio")) +
  geom_point(aes(y = pct_lf, col = "Labor force participation rate")) +
  geom_line(aes(y = pct_lf, color = "Labor force participation rate")) +
  scale_color_manual(values = colors) +
  labs(y = "Percent", x = "", color = "Measure",
       title = "Labor market data: US, 1950–2021") +
  easy_legend_at("bottom") +
  easy_y_axis_title_size(13) +
  easy_plot_legend_size(13) +
  labs(color = "Measure") +
  easy_legend_at("bottom")

```

---

# Basic definitions

We have already talked about how unemployment can be considered a .hi[waste of potential].

  * Both from  *individual* and *aggregate* points of view!
  
--

<br>

When someone is .hi-orange[stuck] in a lower-paid, less productive job, they are considered .hi[underemployed] (*aka* disguisedly unemployed).

--

  * This situation may come about due to periods of *low demand* for jobs in general, or more specifically in a person's area of specialization.
  
  * There is .hi-orange[no] actual *data* on underemployment.



--

.right[Examples?]

---

# Applications

<br>

```{r, echo=FALSE}

ex <- read_csv("example.csv")

ex %>% 
  select(Year, `Adult Population`, Employed, Unemployed) %>% 
  mutate(Year = as.character(Year)) %>% 
  kable(format.args = list(big.mark = ","), caption = "Employment data for the US: 2017–2021") %>% 
  kable_styling(font_size = 22)


```

--


.small[

Calculate the following statistics:

  - (a) The *total labor force*;
  - (b) The number of persons *not in the labor force*;
  - (c) The *unemployment rate*;
  - (d) The *employment-to-population ratio*;
  - (e) The *labor force participation rate*.
]

---


layout: false
class: inverse, middle

# Next time: Unemployment *vs.* other aggregates; types of unemployment


---
exclude: true