---
title: ".b[GDP: Different shapes and forms]"
subtitle: ".b[.hi[EC 103--02]]"
author: "Marcio Santetti"
date: "Fall 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())

```



# Motivation



---

# Housekeeping

 .b[Required readings]:
 

<br>

  * [`OpenStax, ch. 6`](https://openstax.org/books/principles-macroeconomics-2e/pages/6-introduction-to-the-macroeconomic-perspective).

---
layout: false
class: inverse, middle

# Nominal vs. Real


---

# Nominal vs. Real

A .hi[crucial] distinction in Economics is between .hi-orange[real] and .hi-orange[nominal] values.

--

For the case of GDP, this is especially important when .hi[inflation] is such an important component of the economy.

--

> .hi-orange[Inflation] is a *sustained* increase in an economy's price level.

--
  
<br><br>
  
The .hi[nominal] value of any economic measure implies a statistic in terms of .hi-orange[actual prices] that exist at the time. 

--
  
On the other hand, .hi[real] values refer to the *same* statistic after it has been .hi-orange[adjusted for inflation]. 



---

# Nominal vs. Real


Reality check...


<br><br>

[`US Nominal Gross Domestic Product`](https://fred.stlouisfed.org/series/GDP#0)



---

# Nominal vs. Real

Simply looking at *nominal* values when evaluating economic statistics may be .hi[misleading].

--
  
Consider the number of .it[cars sold], for instance.

--
  
  * Does an SUV have the *same* price today, relative to, say, 1985?
  
--
  
<br>
  
  
  
Thus, to have a more .hi-orange[*accurate*] basis for evaluating economic growth, economists prefer .hi[real] measures, as they account for inflation.

--
  
For the case of GDP, we use the .hi-orange[GDP deflator] to .hi[*normalize*] nominal GDP values.

--

  > The .hi[GDP deflator] is a price index measuring the average prices of all final goods and services included in the economy.



---

# Nominal vs. Real

<br>

Before we move on...

--

  * What is a .hi[price index]?

--

An .hi-orange[index] is a number that allows for comparisons across different points in time or different entities.

--

For our purposes, a .hi[price index] is a reference number that allows us to compare economic statistics at different points in time, serving as an *overall average change in relative prices over time*.

--

Over time, the amount of goods and services produced by an economy .hi[increases], and so do .hi-orange[their prices].

--

To that end, the .hi[GDP deflator] is a price index that includes .hi-orange[all] goods and services that are counted in GDP through a *weighted average* methodology.


---

# Nominal vs. Real

Suppose a "toy" economy, with .hi[no] government and .hi[no] interactions with the foreign sector, that only produces .hi[apples].

--

In this case, its .hi-orange[GDP] will simply be the price of apples multiplied by the amount of apples sold.

--

And a few periods:

```{r, echo=FALSE}
toy <- tibble(
  "Year" = c(2015, 2016, 2017),
  "Current Price (per unit)" = c(1.00, 1.10, 1.15),
  "Quantity sold" = c(200, 250, 230),
  "Nominal GDP" = c(200, 275, 264.5)
)  
  
toy %>% 
  kable()


```

<br>

--

In case we want to calculate this economy's .hi[real GDP], we first need to set a .hi-orange[base year].

  - Meaning: we will select a year whose .hi[price level] will be the reference for all years.


---

# Nominal vs. Real

Suppose we set our .hi[base year] to 2016.

--

What we need to do is multiply the .hi[quantities produced] each year by the .hi[base-year price].

--


Then, 


```{r, echo=F}
toy2 <- tibble(
  "Year" = c(2015, 2016, 2017),
  "Constant Price (per unit)" = c(1.10, 1.10, 1.10),
  "Quantity sold" = c(200, 250, 230),
  "Real GDP" = c(220, 275, 253)
)  
  
toy2 %>% 
  kable()
```

---
# Nominal vs. Real

<br><br>

Comparing:


```{r, echo=F}
toy3 <- tibble(
  "Nominal GDP" = c(200, 275, 264.5),
  "Real GDP (Constant 2016 prices)" = c(220, 275, 253)
)  
  
toy3 %>% 
  kable()
```


---

# Nominal vs. Real

Reality is way more complicated than our "toy" economy from the previous example.

--

However, the intuition about the .hi[indexing] procedure remains the same.

--

What the [`Bureau of Economic Analysis (BEA)`](https://www.bea.gov/) does is computing a GDP deflator with the .hi-orange[average prices] of all goods and services included in the economy.

--

<br>

And we obtain real GDP values applying the formula below:

<br>


$$
\begin{aligned}
\text{Real GDP} = \dfrac{\text{Nominal GDP}}{\text{GDP Deflator}}
\end{aligned}
$$

---

# Nominal vs. Real


Reality check...


<br><br><br>


[`The US GDP price deflator`](https://fred.stlouisfed.org/series/A191RI1Q225SBEA)

---

# Nominal vs. Real

Consider the following data (in US$ billions) for the US, between 2008q1 and 2009q4 (base year = 2005):

```{r, echo=F}
real_gdp <- tibble(
  "Quarter" = c("2008q1", "2008q2", "2008q3", "2008q4", "2009q1", "2009q2", "2009q3", "2009q4"),
  "Nominal GDP" = c(14373.9, 14497.8, 14546.7, 14347.3, 14178.0, 14151.2, 14242.1, 14453.8), 
  "Real GDP" = c(13366.9, 13415.3, 13324.6, 13141.9, 12925.4, 12901.5, 12973.0, 13149.5)
)

real_gdp <- real_gdp %>% 
  mutate("GDP deflator" = `Nominal GDP`/`Real GDP`)



real_gdp %>% 
  select(-`Real GDP`) %>% 
  kable(digits = 2)


```

<br>

--

  - Calculate the .hi[Real GDP] for the US economy over this period.


---
layout: false
class: inverse, middle

# Domestic vs. National


---

# Domestic vs. National

A close "cousin" of GDP is the .hi[Gross National Product] (GNP).

--

The main .hi-orange[difference] between the two is that, while GDP only includes what a country produces .hi[within its borders], GNP *adds* what domestic businesses and labor .hi[abroad] produce, and *subtracts* any payments that foreign labor and businesses located in the local country send home to other countries.

--

  - In other words, GNP is based more on what a country's citizens and firms produce, *wherever* they are located, and GDP is based on what happens within a certain country's geographic *boundaries.*
  
--

<br><br>
  
[`US Gross National Product Data`](https://fred.stlouisfed.org/series/GNP)



---
layout: false
class: inverse, middle

# Comparing different countries


---

# Comparing different countries

Applying a similar reasoning to the *Real vs. Nominal* GDP, we can also compare the economic performances among .hi[different countries].

--

To that end, given that almost each nation has its own .hi-orange[currency], we use the .hi[exchange rate] as a common denominator.

  > The .hi-orange[exchange rate] is the value of one currency in terms of *another* currency.


--

<br><br>

Economists typically use the .hi[Purchasing Power Parity (PPP)] measure of GDP to compare the size and performance of different economies.





---

# Comparing different countries

As an example, suppose we would like to compare two countries: Japan and the United States.

--

Suppose that, in 2021, 1 Japanese *Yen* is equivalent to 0.007 US dollar.

  - Thus, the exchange rate Yen/Dollar is 142.85 Yen = US$ 1.
  
--

<br>

If we wish to .hi-orange[convert] Japan's GDP of 541,473.00 Billions of Yen to US dollars, we simply apply the following formula:

<br>

$$
\begin{aligned}
\text{Japan's GDP in US\$} = \dfrac{\text{Japan's GDP in Yen}}{\text{Yen/Dollar Exchange Rate}} = \dfrac{541,473.00}{142.85} = \text{US\$} \ 3,790.50
\end{aligned}
$$

---

# Comparing different countries

<br><br>


In 2021, the Euro/US Dollar exchange rate was 0.98 Euro = 1 US Dollar. 

Also, Germany's GDP was 3,570.62 billions of Euros.

  - Compute Germany's GDP in terms of Purchasing Power Parity (PPP) in US dollars.




---
layout: false
class: inverse, middle

# Next time: Economic growth

---
exclude: true